Let's talk about Thurstone & Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-05-20 paper2.qmd
section:6; process:write

<!-- and spoken language [@Boonen_et_al_2020]. Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_et_al_2015], engineering design skills [@Bartholomew_et_al_2018], build real-time web-based portfolios of performance [@@Kimbell_2012], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020]. -->

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020; @Boonen_et_al_2020].

@@

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003 and @vanDaal_et_al_2016 demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 1. How to derive an stat model from an SCM? -->
With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption in CJ experiments (see @sec-theory-issue1a). This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016]. Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in Equation ([-@eq-mat51]). For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent, where $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
$$
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
$$ {#eq-mat51}

@@

<!-- 2. assumptions about the outcome -->
Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in [SCM -@fig-cj16_scm], the probabilistic model [-@fig-cj16_prob], and the statistical model [-@fig-cj16_stat]. We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Models (GLMs) [@Nelder_et_al_1983; @Nelder_et_al_1996; @Agresti_2015], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.

@@

<!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16. We make this omission because, although these distributions contribute to the overall joint distribution of the data, they do not influence the relationship between $O_{R}$ and $D_{R}$. This independence stems from the random selection designs that govern $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$[^8].

@@

<!-- 4. functional form for D_{R} -->
Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the texts, plus the corresponding repeated judge biases $B_{JK}[j, k]$. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}

This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. In simple terms, the discriminal processes of the stimuli become an observable outcome only through the judges' perceptions (biases). To clarify, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.

@@

<!-- 6. error assumptions -->
Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$. These error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables. Thus, we must resolve these indeterminacies to identify the latent variable model [@Depaoli_2021; @deAyala_2009]. Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix. To address the *location* indeterminacy, we set the errors' mean vector to zero: 
$$
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
$$ {#eq-mat53}

@@

Because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$. We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait. As a result, $T_{IA}$ is also normally distributed. This configuration effectively reinstates Thurstone’s original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).

@@

<!-- How about power?, you have a framework that allows to do this and more. -->

<!-- talk about S and C -->

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge in this model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.

@@

<!-- 3. The issues with Case V, and the BTL model -->
Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b]. In particular, he noted that while assuming equal discriminal dispersions and zero correlation for the stimuli' discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Kelly_et_al_2022; @Andrich_1978]. As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Furthermore, because Thurstone's primary goal was to produce a "coarse scaling" of traits and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the resulting conclusions [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

The study then translates the SCM into a bespoke statistical model that addresses key limitations of the BTL model. This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2). In particular, the model accounts for judge biases, captures the hierarchical structure of stimuli, accommodates heterogeneity in discriminal dispersions, and incorporates measurement error in the hypothesis testing process. These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

<!-- 6. Which are these new avenues for research? -->
Five research avenues merit particular attention for their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect the robustness of statistical inferences. Second, further work is needed to understand how judges' biases impact the number of comparisons required to achieve reliable trait estimates. Third, researchers should re-evaluate existing methods for identifying and excluding 'misfitting' judges and stimuli to ensure their effectiveness. Fourth, studies should explore how the hierarchical structure of stimuli influences both experimental design and hypothesis testing in CJ contexts. Finally, researchers should explore principled approaches for selecting contextually relevant variables.

<!-- 6.2 About the implications of the sampling and comparison mechanisms -->

[About the implications of the sampling and comparison mechanisms]{style="color:blue;"}

<!-- Continuing with the first advantage, now we clarify two additional processes of CJ experiments that previous literature do not typically mention: that the sample and comparison mechanism are generators of missing data--that is, they determine which observations do we get to see in our data sample. This opens the path to use Rubin's missing data framework [@Rubin_1976] into CJ applications to investigate more complex CJ designs, particularly the ones that are not random samples as the ones outlined here, which could be characterized in Rubin's structure parlor as missing at random (MAR) and missing not at random (MNAR) [@Rubin_1976]. Among these cases are adaptive comparative algorithm designs, for which the causal structure is not clear, and thus, its impact on trait measurement and hypothesis testing is also not clear. Another  -->

<!-- normally S and C are not explicitly mentioned  -->

<!-- Remember: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
C -> {Y}
e_{A} -> {C, Y} 
this means that e_{A} is a confounder?,  
then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C} by taking a random sample of stimuli?. Does this mean that random algorithms per se will always be unbiased?. -->

<!-- Brought Rubin's missing data framework to the field. -->
<!-- Because $S$ and $C$, by design, generate data that it is missing completely at random (MCAR) [@Rubin_1976] are random samples, they fulfill ignorability. -->

@@

<!-- power analysis to know the implications of certain sampling is nice to have it explicitly. -->

<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. Previous research has highlighted that factors such as age, culture, and education [@Kelly_et_al_2022, pp. 683], as well as individual differences among judges [@Gill_et_al_2013; @vanDaal_et_al_2017; @vanDaal_2020], could influence judgment accuracy, thereby affecting the "shared consensus" of the trait measurement. As @Kelly_et_al_2022 noted [pp. 683], “Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women?” -->

<!-- Furthermore, it is not inconceivable that the selection of a particular group of judges used to perform the pairwise comparison could result in an incomplete depiction of the dimensions and complexity of the trait. --->

<!-- If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on. Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? [@Kelly_et_al_2022, pp. 683] -->

@@

<!-- 6.3 About judges biases, judgment discrepancies, and the number of judges and judgments -->

[About judges biases, judgment discrepancies, and the number of judges and judgments]{style="color:blue;"}

<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments. Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges. It also accounts for potential judge biases in the comparative judgment process. -->

@@

<!-- 6.1 About the hierarchical structure of stimuli, number and predictors -->

@@

>>>>

<!-- and spoken language [@Boonen_et_al_2020]. Additionally, it has been applied to evaluate various competencies, including mathematical problem-solving skills [@Jones_et_al_2015], engineering design skills [@Bartholomew_et_al_2018], build real-time web-based portfolios of performance [@@Kimbell_2012], conceptual understanding in algebra [@Jones_et_al_2019], statistical and English knowledge [@Marshall_et_al_2020], and STEM knowledge and skills [@Bartholomew_et_al_2020b]. -->

<!-- 2. What is the effectiveness of CJ -->
Numerous studies have documented the effectiveness of CJ in assessing traits and competencies over the past decade. These studies have highlighted three aspects of the method's effectiveness: its reliability, validity, and practical applicability. Research on reliability suggests that CJ requires a relatively modest number of pairwise comparisons [@Verhavert_et_al_2019; @Crompvoets_et_al_2022] to generate trait scores that are as precise and consistent as those generated by other assessment methods [@Coertjens_et_al_2017; @Goossens_et_al_2018; @Bouwer_et_al_2023]. In addition, the evidence suggests that the reliability and time efficiency of CJ are comparable, if not superior, to those of other assessment methods when employing adaptive comparison algorithms [@Pollitt_2012b; @Verhavert_et_al_2022; @Mikhailiuk_et_al_2021]. Meanwhile, research on validity indicates the capacity of CJ scores to represent accurately the traits under measurement [@Whitehouse_2012; @vanDaal_et_al_2016; @Lesterhuis_2018_thesis; @Bartholomew_et_al_2018; @Bouwer_et_al_2023]. Lastly, research on its practical applicability highlights CJ's versatility across both educational and non-educational contexts [@Kimbell_2012; @Jones_et_al_2015; @Bartholomew_et_al_2018; @Jones_et_al_2019; @Marshall_et_al_2020; @Bartholomew_et_al_2020b; @Boonen_et_al_2020].

@@

<!-- 2. The first scenario where the assumption does not hold -->
In the first scenario, research on text quality suggests that when judges evaluate multidimensional, complex traits or heterogeneous stimuli, they often rely on various intricate characteristics of the stimuli to form their judgments [@vanDaal_et_al_2016; @Lesterhuis_2018; @Chambers_et_al_2022]. These additional relevant characteristics, when assessed, are unlikely to be equally weighted or opposing. As a result, they may exert an uneven influence on judges' perceptions, creating biases that resist cancellation. For example, this could occur when a judge assessing the argumentative quality of a text places more weight on its grammatical accuracy than other judges, thereby favoring texts with fewer errors but weaker arguments. Furthermore, since the discriminal difference of the stimuli becomes an observable outcome only through the judges' perceptions, these biases could introduce dependencies between the stimuli [@vanderLinden_et_al_2017_II]. While direct evidence for this particular scenario is lacking, studies such as @Pollitt_et_al_2003, @vanDaal_et_al_2016 and @Bartholomew_et_al_2020a demonstrate the presence of such biases, supporting the notion that the factors influencing pairwise comparisons may not always cancel out.

@@

<!-- 1. How to derive an stat model from an SCM? -->
With the structure in [SCM -@fig-cj15_scm], we can derive a statistical model that addresses violations of the equal dispersion assumption (see @sec-theory-issue1a). This derivation is possible because a fully specified SCM encodes functional and probabilistic information, which we can replace with suitable functions and probabilistic assumptions [@Pearl_et_al_2016]. Specifically, [SCM -@fig-cj15_scm] allows us to express the joint distribution of our complex CJ system as a product of simpler conditional probability distributions (CPDs)[^6], as shown in Equation ([-@eq-mat51]). For clarity, we treat expressions such as $Y := f_{Y}(X)$, $P(Y \mid X)$, and $Y \sim f(Y \mid X)$ as equivalent, where $P(Y \mid X)$ and $f(Y \mid X)$ represent the CPD of $Y$ given $X$.
$$
\begin{aligned}
  P(O_{R}, & S, C, D_{R}, T_{IA}, X_{IA}, e_{IA}, T_{I}, X_{I}, e_{I}, B_{JK}, Z_{JK}, e_{JK}, B_{J}, Z_{J}, e_{J} ) & \\
  &= P(O_{R} \mid D_{R}, S, C) \cdot P(S) \cdot P(C) \cdot P(D_{R} \mid T_{IA}, B_{JK}) \\
  & \quad \cdot P(T_{IA} \mid T_{I}, X_{IA}, e_{IA}) \cdot P(T_{I} \mid X_{I}, e_{I}) \\
  & \quad \cdot P(B_{JK} \mid B_{J}, Z_{JK}, e_{JK}) \cdot P(B_{J} \mid Z_{J}, e_{J}) \\
  & \quad \cdot P(X_{IA}) \cdot P(X_{I}) \cdot P(Z_{JK}) \cdot P(Z_{J}) \\
  & \quad \cdot P(e_{IA}) \cdot P(e_{I}) \cdot P(e_{JK}) \cdot P(e_{J})  
\end{aligned}
$$ {#eq-mat51}

@@

<!-- 2. assumptions about the outcome -->
Each CPD in Equation ([-@eq-mat51]) rests on specific assumptions, which we outline in @fig-cj16. We begin by assuming that $O_{R}$ follows a Bernoulli distribution[^7], reflecting the binary nature of CJ outcomes. Furthermore, following the conventions of Generalized Linear Models (GLMs) [@Nelder_et_al_1983; @Nelder_et_al_1996; @Agresti_2015], the distribution links $O_{R}$ to the latent discriminal difference vector $D_{R}$ using an inverse-logit function: $\text{inv\_logit}(x) = 1/(1 + \exp(-x))$.

@@

<!-- 3. Note about the sample and comparison mechanism, as well as the predictor's distribution -->
Note that while the joint distribution in Equation ([-@eq-mat51]) includes the probability distributions of the sampling and comparison mechanisms, $P(S)$ and $P(C)$, as well as those of the predictor variables--$P(X_{IA})$, $P(X_{I})$, $P(Z_{JK})$, and $P(Z_{J})$--we omit these probabilities from the models presented in @fig-cj16. We make this omission because, although these distributions contribute to the overall joint distribution of the data, the variables  $S$, $C$, $X_{IA}$, $X_{I}$, $Z_{JK}$, and $Z_{J}$ are observed and independent of any other variable in the model. As observed variables, they do not require distributional assumptions--unlike the idiosyncratic errors. Moreover, their independence stems from the random selection designs that govern the variables[^8].

@@

<!-- 4. functional form for D_{R} -->
Next we define $D_{R}$ as the difference between the discriminal processes $T_{IA}[i, a]$ and $T_{IA}[h, b]$, representing the underlying written-quality trait of the compared texts, plus the corresponding repeated judge bias $B_{JK}[j, k]$. Note that if we assume that $B_{JK}[j,k]$ reflects the difference in stimulus-specific biases, i.e., $B_{JK}[j,k] = B_{JK}[i,a,j,k] - B_{JK}[h,b,j,k]$, we can re-write the discriminal difference as: 
$$
\begin{aligned}
D_{R} &= \left( T_{IA}[i,a] - T_{IA}[h,b] \right) + B_{JK}[j,k] \\
&= \left( T_{IA}[i, a] + B_{JK}[i, a, j, k] \right) - \left( T_{IA}[h, b] + B_{JK}[h, b, j, k] \right) \\
& = T^{*}_{IA}[i,a] - T^{*}_{IA}[h,b]
\end{aligned}
$$ {#eq-mat52}

This formulation reveals that the discriminal difference captures a *pure interaction effect*, in which neither the texts' discriminal processes nor the judges' biases alone determine the outcome, but their interaction does [@Attia_et_al_2022]. Put simply, this mathematical description captures the idea that the stimuli' discriminal processes become an observable outcome only through the lens of judges' perceptions (i.e., their biases). For clarity, the square brackets in $D_{R}$ indicate the relevant indices for each trait vector; they do not imply any subsetting of the data.

@@

<!-- 6. error assumptions -->
Next, we specify the probabilistic assumptions for the idiosyncratic errors $e_{I}$, $e_{IA}$, $e_{J}$, and $e_{JK}$. Unlike other variables in the model, these error terms exhibit indeterminacies in their *location*, *orientation*, and *scale* due to the lack of an inherent scale in the associated latent variables $T_{I}$, $T_{IA}$, $B_{J}$, and $B_{JK}$. Thus, to identify the latent variable model, we must resolve these indeterminacies [@Depaoli_2021; @deAyala_2009]. Drawing on principles from SEM [@Hoyle_et_al_2023], we assume that the vector of idiosyncratic errors $\boldsymbol{e} = [e_{I}, e_{IA}, e_{J}, e_{JK}]^{T}$, follows a Multivariate Normal distribution with mean vector $\boldsymbol{\mu}$ and a covariance matrix $\boldsymbol{\Sigma} = \boldsymbol{D} \boldsymbol{R} \boldsymbol{D}$, with $\boldsymbol{D}$ denoting a diagonal matrix of standard deviations and $\boldsymbol{R}$ a correlation matrix. To address the *location* indeterminacy, we set the errors' mean vector to zero: 
$$
\boldsymbol{\mu} = [0, 0, 0, 0]^{T}
$$ {#eq-mat53}

@@

Note that because the error vector $\boldsymbol{e}$ follows an uncorrelated Multivariate Normal distribution, the marginal distribution of $e_{IA}$ is a univariate Normal distribution with mean zero and standard deviation $p_{IA}$. We set $P_{IA}$ as a proportion of $1$ to establish the scale of the stimuli' latent trait relative to the scale of the individuals' trait. As a result, $T_{IA}$ is also normally distributed. This configuration effectively reinstates Thurstone’s original assumption of Normal discriminal processes for the stimuli (see @tbl-thurstone_cases).

@@

<!-- 1. Summary of Thurstone's general theory -->
Thurstone introduced the Law of Comparative Judgment to measure psychological traits of stimuli through pairwise comparisons [@Thurstone_1927a; @Thurstone_1927b]. In its general form, the theory models single-judge comparisons across multiple, potentially correlated stimuli. Each comparison produces a dichotomous outcome indicating which stimulus the judge perceives as having a higher trait level. However, Thurstone identified one key challenge with its measurement model: it required estimating more "unknown" parameters than the number of available pairwise comparisons [@Thurstone_1927b]. To address this issue and facilitate the theory's practical applicability, he formulated five cases derived from this general form, each progressively incorporating several simplifying assumptions.

@@

<!-- 3. The issues with Case V, and the BTL model -->
Although replacing the Normal with the Logistic distribution has a minimal impact on trait estimation or model interpretation [@vanderLinden_et_al_2017_I; @McElreath_2021], Thurstone acknowledged that the two remaining assumptions of Case V--and, by extension, of the BTL model--could pose significant problems when researchers assess complex traits or heterogeneous stimuli [@Thurstone_1927b]. In particular, he noted that while assuming equal discriminal dispersions and zero correlation for discriminal processes simplifies the trait measurement model, these assumptions may fail to reflect the complexity of CJ data [@Thurstone_1927a; @Andrich_1978; @Kelly_et_al_2022]. As a result, they can lead to unreliable and inaccurate trait estimates [@Ackerman_1989; @Zimmerman_1994; @McElreath_2020; @Hoyle_et_al_2023].

Furthermore, because Thurstone aimed primarily to produce a trait's "coarse scaling" and "allocate the compared stimuli on this continuum" [@Thurstone_1927a, pp. 269], his theory did not specify how to use the resulting trait estimates for statistical inference. The CJ tradition has attempted to address this gap by separating trait estimation from the hypothesis testing process. Specifically, CJ studies often rely on point estimates of the traits (i.e., the BTL scores)--or their transformations--to conduct statistical inference. Although separating these processes simplifies the analysis of CJ data, the statistical literature cautions against this practice, as it can introduce bias into the analysis and compromise the reliability of the hypothesis test [@McElreath_2020; @Kline_et_al_2023; @Hoyle_et_al_2023].

@@

The study then translates the SCM into a bespoke statistical model that addresses key limitations of Case V. This model allows researchers to analyze CJ data when the assumptions of equal dispersion and zero correlation do not hold (see @sec-theory-issue1a and @sec-theory-issue1b), and when they need to perform statistical inference (refer to @sec-theory-issue2). In particular, the model accounts for judge biases (see @sec-theory-theoretical_P1), captures the hierarchical structure of stimuli (refer to @sec-theory-theoretical_P3), incorporates measurement error in the hypothesis testing process (refer to @sec-theory-theoretical_P3), and accommodates heterogeneity in discriminal dispersions (see @sec-statistical). These methodological innovations can potentially enhance the reliability and validity of trait measurement [@Perron_et_al_2015] and improve the accuracy of statistical inferences.

@@

<!-- 6. Which are these new avenues for research? -->
Five research avenues deserve particular attention due to their direct impact on the reliability and validity of trait estimates and the accuracy of statistical inferences. First, researchers should investigate how sampling and comparison mechanisms influence CJ experimental outcomes and affect inference robustness. Second, studies should assess the impact of judges' bias on the number of comparisons needed to achieve reliable trait estimates. Third, future work should re-evaluate current methods for identifying 'misfitting' judges and stimuli. Fourth, researchers should analyze how the hierarchical structure of stimuli affects both experimental design and hypothesis testing. Finally, studies should explore how modeling non-dichotomous outcomes or incorporating multidimensional traits affects the reliability and validity of the trait estimates.

<!-- 6.1 About the implications of the sampling and comparison mechanisms -->
The explicit integration of the sampling and comparison mechanisms within our conceptual model offers a novel perspective on the first research avenue. Specifically, it shows how these mechanisms influence the data-generating process by acting as missing data sources, determining which observations are part of the data sample. This new perspective invites the application of Little and Rubin's principled missing data framework [@Little_et_al_2020], enabling a more rigorous evaluation of existing claims about these mechanisms, their influence on CJ outcomes, and their implications for designing and evaluating more complex experimental setups.

<!-- --unlike prior literature, which merely acknowledged their presence in the experimental design. -->

One prominent claim in the CJ literature is that Thurstone's model naturally handles non-random missing data without compromising the reliability or validity of trait estimates [@Bramley_2008]. However, it is easy to imagine scenarios where this assumption may not hold. For instance, selecting a particular group of judges could result in an incomplete representation of the trait's full range and complexity, as judgment accuracy varies with demographic factors [@Bartholomew_et_al_2020a; @Kelly_et_al_2022] and even between individuals [@Gill_et_al_2013; @vanDaal_et_al_2016; @vanDaal_et_al_2017; @vanDaal_2020]. Consequently, judges' selection could affect the formation of a "shared consensus", potentially biasing trait estimates and distorting the observed trait distribution [@Deffner_et_al_2022]. It is therefore essential to determine whether such sampling biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences. Our framework offers a rigorous setting to test the scenario.

[another example: For instance, consider a CJ experiment assessing speech quality using a case-control design [@Everitt_et_al_2010] that samples individuals with normal hearing (NH) and those with hearing impairment (HI). Here, the sampling mechanism depends on hearing status--a variable likely to influence the trait of interest, as HI individuals may exhibit lower speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024]. Suppose a subgroup within the HI population, such as individuals with genetically induced hearing loss, is difficult to reach. If this group is systematically excluded and happens to over- or under-perform in speech quality, the resulting trait estimates may be biased. It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}

[another example: For instance, consider a high-stakes non-mandatory summative CJ assessment, where the results the individuals have in this assessment have some implications for them, however, their participation is not mandatory. In this case, it not hard to imagine that some of these individuals will not participate in the CJ assessment, thus affecting the sample, and ultimately the trait estimates: now a probable under performing group will not available in the data sample, thus biasing the resulting trait estimates. It is therefore essential to determine whether such biases occur--and if so, whether they undermine the validity of trait estimates and the robustness of statistical inferences--and our framework provides a rigorous setting to test this claim.]{style="color:blue;"}

@@

<!-- The logic is that in aggregating the varied judgments, one comes to an overall view of the community of practice. However, this raises important questions about diversity of the judging pool. If it is acknowledged that judges may differ in what they value, it is not inconceivable that judges may differ as a function of their gender, age, ethnicity, culture, class, income, education and so on. Would the aggregate view of young, British-Asian men always be the same as the aggregate view of older, black women? For example, Bartholomew et al. (2020) found differences between what judges from the UK, Sweden, and the USA valued when assessing design portfolios, providing evidence to support divergence along demographic lines. [@Kelly_et_al_2022, pp. 683] -->

@@

Moreover, when designing and evaluating more complex experimental setups, our framework highlights one key principle: the sampling and comparison mechanisms directly influence the CJ outcome. In this study, we deliberately structured these mechanisms to produce data that is *missing completely at random* (MCAR)--that is, the missingness is unrelated to any observed or unobserved variables, including the outcome [@Little_et_al_2020]. This design offers one key advantage: it generates simple random samples that satisfy the condition of ignorability, enabling researchers to legitimately ignore missing data during analysis without introducing bias [@Everitt_et_al_2010; @Kohler_et_al_2019].

However, many modern CJ applications involve more complex forms of missingness--such as *missing at random* (MAR) or *missing not at random* (MNAR) mechanism [@Little_et_al_2020]--that arise from the chosen sampling and comparison methods. Two notable examples are adaptive comparative judgment designs [@Pollitt_2012a; @Pollitt_2012b] and case-control studies [@Everitt_et_al_2010]. In adaptive designs, an algorithm defines the comparison mechanism by dynamically selecting which stimulus pairs judges evaluate. These designs have produced mixed evidence: some studies report improved trait reliability [@Pollitt_et_al_2003; @Pollitt_2012a; @Pollitt_2012b], while others suggest these gains may be artificially inflated [@Bramley_2015; @Bramley_et_al_2019; @Crompvoets_et_al_2020; @Crompvoets_et_al_2022]. In contrast, case-control studies rely on sampling mechanisms linked to non-assignable variables that likely influence the trait of interest. For instance, @Boonen_et_al_2020 selected children based on hearing status, a non-assignable variable that likely affects speech quality [@Boonen_et_al_2023; @Rivera_et_al_2024]. Nevertheless, regardless of their missingness mechanism, both designs could benefit from explicitly articulating their underlying causal structures. Doing so would enable researchers to assess and test their impact on trait estimation and statistical inference. Our framework provides a rigorous foundation for such articulations.

<!-- Comment for myself: when the sampling or comparison mechanism depend on the precision of the measured trait, the DAG can be depicted as: 
{S, C} -> {Y}
e_{A} -> {S, C, Y} 
this means that e_{A} is a confounder?,  
then should you: "control for" the variable?, or the equivalent, break the relationship e_{A} -> {C, S} by taking a random sample of stimuli?. Does this mean that random algorithms per se will always be unbiased?. -->

<!-- 6.2 About judges biases, judgment discrepancies, and the number of judges and judgments -->
Additionally, by explicitly incorporating judges' biases, our framework provides a novel approach to investigate the second research avenue. Specifically, it enables researchers to determine whether such biases occur and, if so, to assess their impact on the reliability and validity of trait estimation. Although some anecdotal evidence already suggests judges' biases exist [@Pollitt_et_al_2003; @vanDaal_et_al_2016; @Bartholomew_et_al_2020a], further research is needed to address several key questions. For instance: Can CJ data be analyzed under the assumption of "sample-free" trait calibration--that is, the estimated trait difference for any two scripts does not depend on which other scripts they are compared with, implying an absence of judges' bias? [@Bramley_2008; @Andrich_1978] If this assumption does not hold, how many judges and judgments are required to ensure acceptable levels of trait reliability and validity? Do training and expertise influence judgment biases, and to what extent? [@Kelly_et_al_2022] Finally, what are the benefits of *repeated measures designs*, in which judges evaluate the same stimulus pairs multiple times? [@Lawson_2015]

@@

<!-- 6.4 About the hierarchical structure of stimuli, number and predictors -->

@@

<!-- Using this principled framework, the study clarifies the hierarchical structure underlying both stimuli and judgments in CJ experiments. Specifically, the model makes explicit how stimuli are nested within individuals and how judgments are nested within judges. -->
